---
title: "Deck of Cards"
author: "Silas Tittes"
date: "July 2, 2015"
output: pdf_document
fontsize: 12pt
---

Yesterday I was day dreaming about probability and decks of cards. Gambling and probability are like old high school pals. One went on to college and got a nice job in science. She plays tennis on the weekends and gives lectures at fine ivy league institutions (among others). The other is a wealthy alcoholic that spends his time in Vegas casinos (among others). They still get together on a rare occasion, but it doesn't usually end well for anyone (card counters). Anyways, I realized it how easy it would be to simulate a deck of cards using (what else?) R.

```{r, echo=FALSE}
library(rmarkdown)
library(knitr)
library(formatR)
opts_chunk$set(tidy=TRUE, tidy.opts=list(width.cutoff=30))
```

Here is the deck of cards:
```{r}
nums <- as.character(2:10) #numbered cards
types <- c("Ace", "Jack", "Queen", "King", nums) #face cards and numbered cards (the ranks)
suites <- c("Hearts","Spades","Diamonds","Clubs") #the suites

#repeat each value in the card types vector by the number of suites, and repeat  each value in the suites vector by the number of types. Then paste the two output vectors together and seperate the items by "_of_" 
deck <- paste(rep(types,length(suites)), rep(suites, length(types)), sep = "_of_") 
```

Did I do it right?
```{r}
52 == length(deck) #hopefully 52 (There's only room for one joker in this blog post.)
52 == length(unique(deck)) #each card is like a snowflake

#split the string back into the suites and types
broken_deck <- unlist(strsplit(deck, split = "_of_")) 

#the  %in% operator is the bomb! 
rep(13, 4) %in% table(broken_deck[1:length(broken_deck) %% 2/3 == 0]) #count of each suite
rep(4, 13) %in% table(broken_deck[1:length(broken_deck) %% 2/3 != 0]) #count of each rank
print(deck, quote = F) #remove the quotes. quotes are for gamblers. 
```

I love R! 4 simple lines and you have a deck of cards to play with. Now with 4 more lines, I'll show you how to write your own solitaire program. No, not really, but maybe one day, with like 4,000 lines at least. However, I will make a few simple card sampling functions for funsies.

```{r}
#function with no input
shuffle <- function( ){
  sample(deck, size = length(deck), replace = F)
}
#same function with input
shuffle <- function( n.cards ){
    sample(deck, size = n.cards, replace = F)  
}
#function with no input
draw <- function( ){
  sample(deck, size = length(deck), replace = T) #notice TRUE not FALSE
}
#same function, different inputs (nothing versus number of samples)
draw <- function( n.cards ){
  sample(deck, size = n.cards, replace = T) #notice TRUE not FALSE
}

```

Here I made two functions, shuffle and draw. Each function is actually two functions with the same name. I could have made all this into one function, but shuffle and draw a different enough conceptually, so I thought I would keep them separate for demonstration purposes. When the shuffle function receives no input, it randomly grabs from the original deck WITHOUT replacement 52 times. Since it grabs randomly, it is in a sense shuffling the card deck. The shuffle function also takes an integer as an input, in which case n number of cards are drawn at random, again without replacement. Draw is the same as shuffle, but it samples WITH replacement. Meaning the same card could be sampled more than once in a series of draws. The two functions really only differ by the letter T and F (for true and false) in the sample function. I should mention this is a simple form of function overloading, where giving input or no input changes the function behavior on the user's end. Since R doesn't make you declare variable types, function overloading with alternative types of variables as inputs doesn't work the same way function overloading works in something like c++. As far as I can tell R instead uses what are called methods (see ?methods and ?setMethods). I won't go into setting methods here (frankly I haven't really tried setting methods myself). Instead we will stick to this simple hack-esh version of overloading that is probably useless in realistic applications. 
Okay, lets shuffle and draw the deck shall we? 

```{r}
#first use of overloaded function returns all the 
shuffle()
shuffle(2)

draw()
draw(2)

#lets see how often we get a particular card on a large number of draws and see how close to 1/52 we get
card <- "Queen_of_Spades"
n.draw <- 100000
exp <- round(1/52 * n.draw,0)
names(exp) <- "counts"
bunch_o_draws <- draw(n.draw)
obs <- sum(card == bunch_o_draws)
print(cbind(obs,exp))
```

Overall, very silly. Hopefully though it demonstrates the ease in which fairly complex tasks can be accomplished very easily in R. I use the sample function daily. Perhaps in a later post I could use the deck of cards for a better application. 

For now I'll leave it at that. Next time, additive alleles and drift! 
